Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-30 Manuscript/manuscript.Rmd
edits

![**Ensemble Framework.** A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline. Cell line specific TAD boundary data was extracted from a Hi-C produced contact matrix of a given resolution. A variety of functional genomic annotations were used to establish the feature space. Each subsequent random forest classification model was trained on an isolated predictor type in conjunction with a particular data re-sampling technique to account for class imbalance. Each model was then validated on a testing set and performances were evaluated using MCC.](Figures/figure3_ensemble_framework_V3.png)

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE FOR CELL LINE SPECIFIC MODELS

We first evaluated the least granular cell line specific random forest classifiers, those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 4A (red)). We then assessing models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 4B (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance.

However, when implemented using distance-type predictors and no re-sampling, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C (red)).

<!--Add K562 results in main paper-->

<!--
Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy and AUC) were less sensitive to models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). 
-->

## SMOTE RE-SAMPLING COUPLED WITH THE INCLUSION OF A DISTANCE-TYPE FEATURE SPACE SUBSTANTIALLY IMPROVED MODEL PERFORMANCE

When assessing performances for random forest classifiers built on balanced data, we found that SMOTE re-sampling provided the optimal MCC consistently outperforming models built on ROS or RUS, across both predictor type and resolution (Figure 4). Specifically, the most improvement in model performance was seen when combining SMOTE re-sampling with distance-type predictors. This was also the optimal combination for the K562 cell line (Supplementary Figure 3). Interestingly, we found that there was a degredation in model performance as data resolution decreased. Chromosome specific models built on 10 kb data resolutions presented with the highest MCC values. Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics. These results imply that an optimally performing random forest classifier, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.

![**SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/figure4_gm12878_model_performances.png)

>>>>

![**Ensemble Framework.** A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline. Cell line specific TAD boundary data was extracted from a Hi-C produced contact matrix of a given resolution. A variety of functional genomic annotations were used to establish the feature space. Each subsequent random forest classification model was trained on an isolated predictor type in conjunction with a particular data re-sampling technique to account for class imbalance. Ten-fold cross validation was performed to tune for the number of predictor-type specific genomic features to consider at each node. Each model was then validated on a testing set and performances were evaluated using MCC.](Figures/figure3_ensemble_framework_V3.png)

## SMOTE RE-SAMPLING COUPLED WITH A DISTANCE-TYPE FEATURE SPACE PROVIDED OPTIMAL CELL LINE SPECIFIC MODEL PERFORMANCE

<!--## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE FOR CELL LINE SPECIFIC MODELS-->

We first evaluated the least granular cell line specific random forest classifiers, those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 5 A-B (red)). We then assessed models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 5 C-D (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance, regardless of cell line and resolution. However, when cell line specific classifiers were implemented using distance-type predictors and no re-sampling, we found that there was a significant improvement in MCC compared to the other two predictor types (Figure 5 E-F (red)).

We then further explored the impact of data re-sampling on model performance, across cell line and resolution. For cell line specific random forest classifiers built on balanced data, we found that SMOTE re-sampling provided the optimal MCC, consistently outperforming models built on ROS or RUS, across both predictor type and resolution (Figure 5). Specifically, the most improvement was seen when combining SMOTE re-sampling with distance-type predictors for both the GM12878 and K562 cell lines.

## MODEL PERFORMANCE DECREASED AS Hi-C DATA RESOLUTION DECREASED

Interestingly, we found that there was a degredation in model performance as data resolution decreased. Chromosome specific models built on 10 kb data resolutions presented the highest MCC values (Figure 5). Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics. These results imply that an optimally performing cell line specific random forest classifier <!--, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries,--> is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.

![**SMOTE re-sampling coupled with distance-type predictors yielded optimal MCCs across resolution for cell line specific models.** Performances of random forest classification models built on each of the 3 predictor types includeing (A-B) OC, (C-D) OP, and (E-F) Distance, and across both the GM12878 and K562 cell lines. Within each cell line, predictor type, and resolution, chromosome-specific training data was re-sampled using different techniques including no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and *Synthetic Minority Over Sampling TEchnique* (SMOTE; purple). Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v2.png)
